---
title: "Take-Home_Ex01 part 2"
author: "Teo Wee Siang Roy"
date: "May 08, 2025"
date-modified: "last-modified"
format:
  html:
    toc: true
    theme: default
    code-fold: show
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
freeze: true
---
# Reviewed Submission

This review is based on the submission by **Sandra Jacob**, available at:  
🔗 [https://sandrajacob-isss608.netlify.app/takehome_exercises/takehome_ex01/takehomeex01](https://sandrajacob-isss608.netlify.app/takehome_exercises/takehome_ex01/takehomeex01)

# Load packages
```{r}
pacman::p_load(tidyverse, haven,
               ggrepel, ggthemes,
               ggridges, ggdist,colorspace,ggdist,
               patchwork, scales, ggplot2, 
               matrixStats, treemapify, showtext)
```

# Load dataset and data processing
```{r}

respopagesex2024 <- read_csv("TH01_2/data/respopagesex2024.csv")
show_col_types = FALSE

knitr::kable(head(respopagesex2024))
```

```{r}
respopagesex2024[duplicated(respopagesex2024),]
```

```{r}
planning_areas <- read_csv("TH01_2/data/Planning_areas_of_Singapore_2.csv")
```

```{r}
respop_with_region <- respopagesex2024 %>%
  left_join(planning_areas, by = c("PA" = "Name (English)"))

respop_cleaned <- respop_with_region %>%
  select(-`Time`, -`Malay`, -`Chinese`, -`Pinyin`, -`Tamil`, -`Area (km2)`, -`Population`, -`Density (/km2)`, -`Subzones`)
```

# Gender Ratio Analysis by Region (As reference)

## ✅ Three Good Design Principles Applied

1. **Clear and Descriptive Title with Proper Axis Labels**
The chart’s title — *"Male and Female Population by Region"* — succinctly conveys the purpose of the visualisation. Coupled with clearly labeled axes (*"Region"* and *"Total Population"*), this design adheres to fundamental **graphical integrity** and supports rapid comprehension. It aligns with best practices for enhancing **cognitive accessibility** in data graphics.

2. **Strategic Use of Color for Categorical Differentiation**
The use of high-contrast, intuitive colors (maroon for Females and navy for Males) effectively encodes the **categorical variable (Sex)**. This leverages **pre-attentive visual processing**, enabling users to instantly distinguish between gender groups. The visual encoding is both **aesthetically pleasing and functionally efficient**, which is crucial in comparative visualisations.

3. **Inclusion of Percentage Labels to Add Analytical Depth**
Each bar is annotated with percentage labels that represent the proportion of each gender within its respective region. This additional layer of information transforms a basic bar chart into a more **analytically enriched visual**, allowing users to compare both absolute values and relative gender compositions simultaneously. This reflects the principle of **"deriving insight" rather than merely displaying raw data"** — a key goal in visual analytics.


## ❌ Three Areas for Further Improvement

1. **Label Overlap and Legibility Concerns**

While the inclusion of percentage labels is commendable, their placement **above the bars** results in visual congestion—especially in regions like *North-East* and *West*, where the height differences between segments are minimal. This diminishes readability and increases cognitive load. To resolve this:

- Place labels **inside the bars** using `position_stack(vjust = 0.5)`
- Use a **horizontal layout** to increase label spacing
- Reduce font size or apply `ggrepel` for smart label adjustment

2. **Lack of Alignment Between Data and Message**

Although percentage labels are presented, the **bar height still encodes absolute values**. 
This can lead viewers to focus on population differences rather than proportional differences. 
A **100% stacked bar chart** would better align with the intent to highlight **gender balance within each region**.

3. **Suboptimal Chart Type for Comparative Ratios**

Grouped bar charts are ideal for comparing totals but are less effective for showing **composition within a group**. 
A **normalized horizontal stacked bar chart** would:

- More clearly show **gender proportion**
- Eliminate vertical label collisions
- Follow a natural **left-to-right reading flow** that supports comparison

## Graph as reference first graph 
```{r}
gender_summary_long <- respop_cleaned %>%
  group_by(Region, Sex) %>%
  summarise(Total_Pop = sum(Pop), .groups = "drop")

#Calculate total population per region
gender_summary_long <- gender_summary_long %>%
  group_by(Region) %>%
  mutate(
    Region_Total = sum(Total_Pop),
    Percent = (Total_Pop / Region_Total) * 100
  )

#Plot side-by-side bars
ggplot(gender_summary_long, aes(x = Region, y = Total_Pop, fill = Sex)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = paste0(round(Percent, 1), "%")),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3, family = "nunito"
  ) +
  labs(
    title = "Male and Female Population by Region",
    x = "Region",
    y = "Total Population"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_family = "nunito") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5),plot.title = element_text(hjust = 0.5, size = 15, face = "bold", family = "nunito")) +
  scale_fill_manual(values = c("Females" = "#662633", "Males" = "#2C3C60"))
```
## 🔧 Improvement 1: What Was Improved

```{r,fig.width= 10,fig.height= 6}
gender_summary_long <- respop_cleaned %>%
  group_by(Region, Sex) %>%
  summarise(Total_Pop = sum(Pop), .groups = "drop") %>%
  group_by(Region) %>%
  mutate(
    Region_Total = sum(Total_Pop),
    Percent = (Total_Pop / Region_Total) * 100,
    Label = paste0(round(Percent, 1), "%\n(", scales::comma(Total_Pop), ")")
  )

# Plot: stacked bar with % + population inside bars
ggplot(gender_summary_long, aes(x = Region, y = Percent, fill = Sex)) +
  geom_col(position = "stack", width = 0.6) +

  # Percentage + population inside each segment
  geom_text(
    aes(label = Label),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 3, family = "nunito"
  ) +

  labs(
    title = "Gender Distribution by Region (2024)",
    subtitle = "Each bar shows gender % and absolute population count",
    x = "Region",
    y = "Percentage (%)",
    fill = "Gender"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = c("Females" = "#A93226", "Males" = "#2C3E50")) +
  theme_minimal(base_family = "nunito") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```

This revised visualisation enhance both analytical clarity and graphical integrity:

### 1. From Absolute Counts to Proportional Comparison
- **Original**: Displayed only raw population counts for each gender, which made it difficult to compare the gender distribution within each region.
- **Improved**: Transformed into a **100% stacked bar chart** that normalises each region to 100%, making it much easier to compare **gender ratios** across regions.

### 2. Embedded Dual-Level Information in Labels
- **Original**: Included either percentage or total population as labels, but not both.
- **Improved**: Combined both **percentage and absolute count** in each label (e.g., `48.7%\n(123,456)`), allowing the viewer to simultaneously interpret **relative proportion and actual magnitude**.

### 3. Improved Label Positioning and Readability
- **Original**: Percentage labels were placed outside the bars, leading to overlapping and reduced clarity, especially for regions with similar male/female proportions.
- **Improved**: Labels are now **placed inside the bar segments** using `position_stack(vjust = 0.5)`, increasing legibility and reducing visual clutter.

### 4. Chart Type Aligns with Analytical Purpose
- **Original**: Used a grouped bar chart, which is more appropriate for comparing absolute values across categories.
- **Improved**: Switched to a **stacked bar layout**, which better supports the analytical objective of comparing **gender balance within each region**.

### Improvement outlines
These improvements follow the visual design principles outlined in 
*Lesson 2: Designing Graphs to Enlighten*, particularly:

- Using the **right chart to match the message**
- Enhancing **interpretability** through derived metrics
- Leveraging **pre-attentive features** like color and label placement

## 🔧 Improvement Summary 2: What Was Further Improved

```{r,fig.width= 10,fig.height= 6}
gender_summary_long <- respop_cleaned %>%
  group_by(Region, Sex) %>%
  summarise(Total_Pop = sum(Pop), .groups = "drop") %>%
  group_by(Region) %>%
  mutate(
    Region_Total = sum(Total_Pop),
    Region_Percent = round((Total_Pop / Region_Total) * 100, 1),
    Label = paste0(Region_Percent, "%\n(", scales::comma(Total_Pop), ")")
  ) %>%
  ungroup() %>%
  mutate(
    Region = fct_reorder(Region, Region_Total, .desc = TRUE),  # sort by Region_Total
    Sex = factor(Sex, levels = c("Males", "Females"))           # ensure stacking order
  ) %>%
  arrange(Region, Sex)  # ensure correct stacking order

# Plot
ggplot(gender_summary_long, aes(y = Region, x = Total_Pop, fill = Sex)) +
  geom_col(position = "stack", width = 0.6) +
  geom_text(
    aes(label = Label),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 3,
    family = "nunito"
  ) +
  labs(
    title = "Total Male and Female Population by Region (2024)",
    subtitle = "Labels show % within each region and absolute population",
    x = "Total Population",
    y = "Region",
    fill = "Gender"
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    breaks = seq(0, 1000000, 100000)
  ) +
  scale_fill_manual(values = c("Females" = "#A93226", "Males" = "#2C3E50")) +
  theme_minimal(base_family = "nunito") +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )
```
Building on the earlier stacked percentage chart, this updated visualisation improves the clarity, comparability, and contextual storytelling of the gender distribution across regions in 2024.

### 1. Switched from Percentage to Actual Population on the X-Axis
- **Previous**: X-axis showed gender **percentage share**, which helped with proportion comparison but hid the actual population scale.
- **Improved**: Uses **raw population values** on the x-axis (with comma formatting and 100k breaks), enabling clear comparison of **both gender ratio and total population** per region.

### 2. Horizontal Layout Enhances Label Readability and Comparison
- **Previous**: Vertical bars made category names harder to read, and labels could overlap in crowded segments.
- **Improved**: **Horizontal stacked bars** make region names easier to scan and reduce label collision, improving overall **legibility** and alignment with **natural left-to-right reading flow**.

### 3. Regions Sorted by Total Population
- **Previous**: Regions were presented in arbitrary or default order, which reduced analytical clarity.
- **Improved**: Regions are now **sorted in descending order of total population**, highlighting population scale differences and making the chart easier to interpret at a glance.

### 4. Maintained Rich, Dual-Layer Labels
- Each bar segment still includes a **concise label** that combines the percentage share and absolute count (e.g., `50.5% (302,710)`), enabling multi-level insight in a compact, readable form.


These refinements not only retain the strengths of the previous version but enhance the visual’s ability to communicate both **structure (proportions)** and **magnitude (totals)** — an important principle covered in *Lesson 2: Designing Graphs to Enlighten*.
