---
title: "Hands-on Exercise 8"
author: "Teo Wee Siang Roy"
date: "May 20, 2025"
date-modified: "last-modified"
format:
  html:
    toc: true
    theme: default
    code-fold: show
editor: visual
execute:
  eval: true
  echo: true
  warning: true
  message: true
freeze: true
---

# Choropleth Mapping with R

## Use package
```{r}
pacman::p_load(sf, tmap, tidyverse)

```

## Import dataset 
```{r}
mpsz <- st_read(dsn = "chap08/data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
mpsz

```


## Importing Attribute Data into R
```{r}
popdata <- read_csv("chap08/data/aspatial/respopagesextod2011to2020.csv")


```
## Data preparation
```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

## Joining the attribute data and geospatial data

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)

mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))

write_rds(mpsz_pop2020, "chap08/data/rds/mpszpop2020.rds")
```


# Choropleth Mapping Geospatial Data Using tmap

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")

```

##  Creating a choropleth map by using tmap’s elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))

```

##  Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```


## Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

##  Drawing a choropleth map using tm_fill() and *tm_border()**

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```


```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,  
             fill_alpha = 0.1)
```

## Data classification methods of tmap
### Plotting choropleth maps with built-in classification methods

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}

tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

### Plotting choropleth map with custome break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```
```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme
```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### Map Legend
```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

### Map style
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture
```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

```{r}
tmap_style("white")
```

## Drawing Small Multiple Choropleth Maps
```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```
```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

##  By defining a group-by variable in tm_facets()
```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

## By creating multiple stand-alone maps with tmap_arrange()
```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

# Mappping Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill(
    "DEPENDENCY",
    style = "quantile",
    palette = "Blues",
    legend.hist = TRUE,
    legend.hist.z = 0.2,
    labels = c("Low", "Medium", "High") 
  ) +
  tm_layout(
    legend.outside = TRUE,
    legend.position = c(0.8, 0.2), 
    legend.width = 1.5,
    legend.height = 0.4,
    legend.text.size = 1.2,
    legend.title.size = 1.4,
    outer.margins = c(0.05, 0.05, 0.05, 0.05),
    frame = FALSE
  ) +
  tm_borders(alpha = 0.5)
```

#import Data 
```{r}
sgpools <- read_csv("chap08/data/aspatial/SGPools_svy21.csv")
```

```{r}
list(sgpools) 

```

## Creating a sf data frame from an aspatial data frame
```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

```{r}
list(sgpools_sf)

```

## Drawing Proportional Symbol Map
```{r}
tmap_mode("view")

```

## It all started with an interactive point symbol map

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```


## Proportional
```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)

```


## Different colour

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```


## Side by side
```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```


```{r}
tmap_mode("plot")



```