---
title: "Hands-on_Ex03"
author: "Teo Wee Siang Roy"
date: "April 29, 2025"
date-modified: "last-modified"
format: html
editor: visual
execute: 
  eval: true
  echo: true
  warning: false
freeze: true
---

# 3 Programming Interactive Data Visualisation with R

```{r}
pacman::p_load(ggiraph, plotly,patchwork, DT, tidyverse) 
```

## Importing the data

```{r}
exam_data <- read_csv("chap03/data/Exam_data.csv")
``` 

```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

### Enhancing Dot Interactivity for Clearer Hover Feedback

To make it easier for users to identify which dot they are hovering over, we enhance the dot's appearance by changing its size and color during hover.

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 
theme_set(theme_minimal(base_size = 14, base_family = "Arial"))

p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = tooltip, data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot",
    fill = "#2c7fb8",        # Soft blue fill
    color = "white",         # Dot border
    alpha = 0.85
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Distribution of Maths Scores",
    subtitle = "Each dot represents a student. Hover to see details.",
    caption = "Source: Exam_data.csv",
    x = "Maths Score",
    y = NULL
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5),
    axis.text.x = element_text(size = 12),
    panel.grid.major.y = element_blank()
  )


options(ggiraph.tooltip_extra_css = "pointer-events: none;")

# Create the interactive widget
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8 * 0.618,
  options = list(
    opts_tooltip(css = "color: red; font-weight: bold; background-color: #f9f9f9;
                         border: 1px solid #ccc; padding: 6px 10px;
                         border-radius: 6px; box-shadow: 2px 2px 5px rgba(0,0,0,0.1);"),
    opts_hover(css = "fill:#f03b20;stroke:white;stroke-width:1.5px;")
  )
)
```

## 3.6.2 Displaying statistics on tooltip

```{r}

tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

### How to improve 

```{r}
exam_data <- exam_data %>%
  mutate(RACE = fct_reorder(RACE, MATHS, .fun = mean, .desc = FALSE))

tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste0("Mean Maths Score: ", mean, "\n± Standard Error: ", sem)
}

gg_point <- ggplot(data = exam_data, aes(x = RACE)) +
  stat_summary(
    aes(y = MATHS, tooltip = after_stat(tooltip(y, ymax))),
    fun.data = "mean_se",
    geom = GeomInteractiveCol,
    width = 0.5,
    fill = "lightblue"
  ) +
  stat_summary(
    aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.2,
    size = 0.2
  ) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  labs(
    title = "Average Maths Scores by Race",
    x = "Race",
    y = "Maths Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14)
  )

girafe(
  ggobj = gg_point,
  width_svg = 8,
  height_svg = 8 * 0.618,
  options = list(
    opts_hover(css = "fill:orange;stroke:black;stroke-width:1.5px;"),
    opts_tooltip(css = "background-color:white; color:black; border:1px solid black; padding:5px; border-radius:5px;")
  )
)

```
## 3.6.5 Combining tooltip and hover effect
```{r}
p <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

### Improve to show additional stats of the class scores
```{r}
exam_data <- read_csv("chap03/data/Exam_data.csv")
class_summary <- exam_data %>%
  group_by(CLASS) %>%
  summarise(
    mean_score = round(mean(MATHS), 2),
    min_score = min(MATHS, na.rm = TRUE),
    max_score = max(MATHS, na.rm = TRUE),
    count = n(),
    .groups = "drop"
)

# Join the stats back to main data and add tooltip content
exam_data <- exam_data %>%
  left_join(class_summary, by = "CLASS") %>%
  mutate(tooltip_text = paste0(
    "Class: ", CLASS, "\n",
    "Student Score: ", MATHS, "\n",
    "Class Mean: ", mean_score, "\n",
    "Min Score: ", min_score, "\n",
    "Max Score: ", max_score, "\n",
    "Count: ", count
  ))

# Dotplot with enhanced tooltip
p <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = tooltip_text, data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot",
    fill = "#1f77b4",
    color = "white",
    alpha = 0.85
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Distribution of Maths Scores",
    x = "Maths Score"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.text.x = element_text(size = 12)
  )

# Interactive rendering
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618,
  options = list(
    opts_hover(css = "fill: #202020; stroke: white; stroke-width: 1.2px;"),
    opts_hover_inv(css = "opacity: 0.15;"),
    opts_tooltip(css = "background-color: white; color: black; border: 1px solid #ccc; padding: 5px; border-radius: 4px; font-size: 12px;")
  )
)
```

##3.6.7 Coordinated Multiple Views with ggiraph
```{r}
p1 <- ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```
### try out how to display all 3 subjects
```{r, fig.width=12, fig.height=20}
exam_data <- exam_data %>%
  mutate(
    total_score = MATHS + ENGLISH + SCIENCE,
    tooltip_text = paste0(
      "Maths: ", MATHS, "\n",
      "English: ", ENGLISH, "\n",
      "Science: ", SCIENCE, "\n",
      "Total: ", total_score, " / 300"
    )
  )

# Step 2: Maths Plot
p1 <- ggplot(data = exam_data, aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID, tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot",
    fill = "#1f77b4",
    color = "white",
    alpha = 0.85
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Maths Score", x = "MATHS") +
  theme_minimal()

# Step 3: English Plot
p2 <- ggplot(data = exam_data, aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID, tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot",
    fill = "#2ca02c",
    color = "white",
    alpha = 0.85
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "English Score", x = "ENGLISH") +
  theme_minimal()

# Step 4: Science Plot
p3 <- ggplot(data = exam_data, aes(x = SCIENCE)) +
  geom_dotplot_interactive(
    aes(data_id = ID, tooltip = tooltip_text),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot",
    fill = "#ff7f0e",
    color = "white",
    alpha = 0.85
  ) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(title = "Science Score", x = "SCIENCE") +
  theme_minimal()

# Step 5: Combine the 3 plots vertically
combined_plot <- p1 / p2 / p3

# Step 6: Render as interactive graphic
girafe(
  code = print(combined_plot),
  width_svg = 9,
  height_svg = 5,
  options = list(
    opts_hover(css = "fill: red; stroke: black; stroke-width: 1.5px;"),
    opts_hover_inv(css = "opacity: 0.2;"),
    opts_tooltip(css = "background-color: white; color: black; border: 1px solid #ccc; padding: 6px; border-radius: 5px; font-size: 12px;")
  )
)
```