---
title: "Hands-on Exercise 2"
author: "Teo Wee Siang Roy"
---

# Getting Started

I will first work through the exercise to gain a clear understanding of the data and objectives. Based on this foundation, 
I will then explore opportunities to apply visual analytics techniques to uncover insights and enhance data storytelling.

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.


```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, hrbrthemes, patchwork)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Following as per given session 2 chapter
## Trying out ggrepel

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Beyond ggplot theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```