---
title: "Hands-on Exercise 2"
author: "Teo Wee Siang Roy"
---

# Getting Started

I will first work through the exercise to gain a clear understanding of the data and objectives. Based on this foundation, 
I will then explore opportunities to apply visual analytics techniques to uncover insights and enhance data storytelling.

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.


```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, hrbrthemes, patchwork)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Following as per given session 2 chapter
## Trying out ggrepel

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Beyond ggplot theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_economist() +
  ggtitle("Distribution of Maths scores") 
```

## Beyond ggplot theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

##  Beyond Single Graph

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  # Replace 'size' with 'linewidth'
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3

````


##  Combining two ggplot2 graphs

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p1 + p2

````


##  Combining two ggplot2 graphs

```{r}

p1 + p2 | p3

````
##  Combining two ggplot2 graphs

```{r}

((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')

````

##  Combining two ggplot2 graphs

```{r}

p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)

````

##  Combining two ggplot2 graphs

```{r}

patchwork <- (p1 / p2) | p3
patchwork & theme_economist()

````