---
title: "Hands-on Exercise 2"
author: "Teo Wee Siang Roy"
---

# Getting Started

I will first work through the exercise to gain a clear understanding of the data and objectives. Based on this foundation, I will then explore opportunities to apply visual analytics techniques to uncover insights and enhance data storytelling.

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse, ggrepel, ggthemes, hrbrthemes, patchwork, ggtern, ggplot2)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Trying out ex 2

```{r}
english_histogram <- ggplot(exam_data, aes(x = ENGLISH)) + 
  geom_histogram(bins = 20, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Distribution of English Scores", 
       x = "English Score", 
       y = "Frequency") + 
  theme_minimal()

math_histogram <- ggplot(exam_data, aes(x = MATHS)) + 
  geom_histogram(bins = 20, fill = "lightgreen", color = "black", alpha = 0.7) + 
  labs(title = "Distribution of Math Scores", 
       x = "Math Score", 
       y = "Frequency") + 
  theme_minimal()

science_histogram <- ggplot(exam_data, aes(x = SCIENCE)) + 
  geom_histogram(bins = 20, fill = "salmon", color = "black", alpha = 0.7) + 
  labs(title = "Distribution of Science Scores", 
       x = "Science Score", 
       y = "Frequency") + 
  theme_minimal()

combined_plot <- (english_histogram / math_histogram / science_histogram) + 
  plot_layout(heights = c(2, 2, 2))

combined_plot
```

## Data Comparison: Class and Subject

```{r}
exam_data_long <- exam_data %>%
  pivot_longer(cols = c(ENGLISH, MATHS, SCIENCE), names_to = "Subject", values_to = "Score")

score_comparison_plot <- ggplot(exam_data_long, aes(x = CLASS, y = Score, fill = Subject)) +
  geom_boxplot() + 
  facet_wrap(~ Subject, scales = "free_y") + 
  labs(title = "Comparison of Scores by Class and Subject",
       x = "Class", y = "Scores") +
  theme_linedraw() + 
  coord_cartesian(ylim = c(0, 100)) +
  theme(legend.position = "top")

score_comparison_plot
```

## Density Plot of Average Scores

```{r}
exam_data$average_score <- (exam_data$ENGLISH + exam_data$MATHS + exam_data$SCIENCE) / 3

ggplot(exam_data, aes(x = average_score)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Average Scores", x = "Average Score", y = "Density") +
  theme_minimal()

```

## density plot by gender

```{r}

ggplot(exam_data, aes(x = average_score, fill = GENDER)) +
  geom_density(alpha = 0.5) +  # alpha makes it semi-transparent
  labs(title = "Density Plot of Average Scores by Gender", x = "Average Score", y = "Density") +
  theme_minimal()


```

## Try out Patchwork

```{r}

histogram_plot <- ggplot(exam_data, aes(x = ENGLISH)) + 
  geom_histogram(bins = 20, fill = "lightblue", color = "black", alpha = 0.7) + 
  labs(title = "Distribution of English Scores", x = "English Score", y = "Frequency") + 
  theme_minimal()

# 2. Scatter plot of MATHS vs SCIENCE scores
scatter_plot <- ggplot(exam_data, aes(x = MATHS, y = SCIENCE)) + 
  geom_point(color = "purple") + 
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Maths vs Science Scores", x = "Math Score", y = "Science Score") +
  theme_minimal()

# 3. Violin plot of average scores by class
exam_data$average_score <- (exam_data$ENGLISH + exam_data$MATHS + exam_data$SCIENCE) / 3
violin_plot <- ggplot(exam_data, aes(x = CLASS, y = average_score, fill = CLASS)) + 
  geom_violin() +
  labs(title = "Average Scores by Class", x = "Class", y = "Average Score") + 
  theme_minimal()

# Combine the plots using patchwork
combined_plot <- (histogram_plot | scatter_plot) / violin_plot

# Display the combined plot
combined_plot


```

# As per given code in Ex2

## Beyond ggplot theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_economist() +
  ggtitle("Distribution of Maths scores") 
```

## Beyond ggplot theme

```{r}
ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## Beyond Single Graph

```{r}
p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  # Replace 'size' with 'linewidth'
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

p3

```

## Combining two ggplot2 graphs

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p1 + p2

```

## Combining two ggplot2 graphs

```{r,  fig.width= 12 , fig.height= 7}

p1 + p2 | p3


```

## Combining two ggplot2 graphs

```{r,  fig.width= 12 , fig.height= 7}

((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')

```

## Combining two ggplot2 graphs

```{r}

p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)

```

## Combining two ggplot2 graphs

```{r,  fig.width= 12 , fig.height= 7}

patchwork <- (p1 / p2) | p3
patchwork & theme_economist()

```
